DynaProg for Scala

### 2013-01-14 docs/report/design.tex
Rework of intro.

\newpage
\section{Architecture design and technical decisions}
% ------------------------------------------------------------------------------------------------
\subsection{User facing language requirements} \label{user_lang}
The field of dynamic programming has been influenced in the recent years by a methodology known as Algebraic Dynamic Programming which uses a grammar and an algebra to separate between the parsing and the score computation:
\quote{The Algebraic Dynamic Programming approach (ADP) introduces a conceptual splitting of a DP algorithm into a recognition and an evaluation phase. The evaluation phase is specified by an evaluation algebra, the recognition phase by a yield grammar. Each grammar can be combined with a variety of algebras to solve different but related problems, for which heretofore DP recurrences had to be developed independently. Grammar and algebra together describe a DP algorithm on a high level of abstraction, supporting the development of ideas and the comparison of algorithms.}

@@

% ------------------------------------------------------------------------------------------------
\newpage
\subsection{Parsing grammar (ADP)}
In this section, we provide a concise description of the Algebraic Dynamic Programming. Would the reader be interested in the mathematical details of the notation, we encourage him to read the section 3 of \cite{adp}.

@@

% ------------------------------------------------------------------------------------------------
%\newpage
\subsection{Recurrences analysis}
\subsubsection{Dead rules elimination}
Dead rules\footnote{\textit{Rule} denotes a tabulation belonging to the grammar; both terms refer to the same concept, with the subtle difference that \textit{rule} emphasizes its grammar membership.} elimination analysis is straightforward: starting from the grammar's axiom, recursively collect all tabulations involved in the computation in $R$ (set of rules that are mandatory in the grammar). The dead rules $D=S \setminus R$ (where $S$ is the set of all tabulations) can safely be removed from the grammar rules. Although seemingly useless for the Scala implementation, this step is necessary to maintain coherency between Scala and CUDA rules numbering (that happen in a later stage on the valid rules). In CUDA, this analysis not only provides dead code elimination, but it also prevents useless computation execution, since all rules are computed sequentially for a particular subsequence before the next subsequence is processed.

\subsubsection{Yield analysis}
Since the original ADP introduces many concatenation combinators\footnote{ADP's original combinators are: $\sim\sim, \sim\sim*, *\sim\sim, *\sim*, -\sim\sim, \sim\sim-, +\sim\sim, \sim\sim+, +\sim+$.} to differentiate empty/non-empty, and floating/fixed-length concatenations, it is quite involved for the programmer to make sure that the concatenation operators exactly match the size of each pair of subsequence involved. Additionally the priority of operators varies in Scala, depending the operator's first character whereas it is possible to specify arbitrary priorities in Haskell. To overcome these issues, we propose to automate the computation of the minimum/maximum length of (subsequences) parsers.

@@

% ------------------------------------------------------------------------------------------------
\subsection{Backtracking}
In order to produce an efficient transformation from ADP-like problem description to plain C recurrences, we need to construct bottom-up recurrences from top-down parser rules. To do that, we slightly need to modify the ADP parsers in order to separate the backtracking and the scoring, because we want to obtain an efficient algorithm: backtrack writes are in $O(n^2)$ whereas score reads are proportional to the algorithmic complexity ($O(n^3)$ or more for non-serial). To deal with this problem, we are facing two options:\ul
\item \textbf{Explicit backtracking:} requires clear syntactical separation between the score and the backtrack which is not implemented in ADP, unless we consider the whole backtrack being part of the scoring (which has a big performance impact and non-constant memory requirement issues that make such GPU implementation hard and not desirable). Additionally, since the backtracking data is user-defined, there is no way to generate the backtracking algorithm automatically, hence the user also needs to provide it.
\item \textbf{Implicit backtracking:} implies that every rule needs to be normalized, and transformed such that given a rule identifier and a set of indices (subproblems breaking), it is possible to retrieve the sub-solutions combination that contribute to the problem solution. To do that we need to apply the following transformations\ol
	\item Normalize rules and identify them uniquely by exploding alternatives: each rule is decomposed into the union of multiple sub-rules uniquely identified by an index, where sub-rules do not contain alternatives (Or parsers). Let $s$ a subrule and $r_s$ its identifier, we also establish a mapping $T$ from identifier to subrule: $(r_s \to s) \in T$.
	\item Let ${\rm cc}(r_s)$ the number of concatenation contained in the sub-rule $r_s$. The data element corresponding to a rule is a pair (score, backtrack) and is named after the tabulation.\ul
		\item The score part consists of a user-defined type (a composite of primitive types, case classes and tuples)
		\item The backtrack part is a tuple $(r_s,(k_1, k_2, \ldots , k_m))$ where $m$ is the maximal number of concatenations occurring in the enclosing rule of $r_s$; more formally $m=\max_z\big[{\rm cc}(r_z) | r_z \in {\rm rule}(r_s)\big]$, and let $m_s = {\rm cc}(r_s) \le m$.
	\ule
	\item During the matrix computation of cell $(i,j)$, if the sub-rule $r_s$ applies, the backtrack will be set as $(r_s,(k_1,k_2,\ldots,k_{m_s}))$; with $i\le k_1\le k_2\le \ldots k_{m_s}\le j$. Note that if the backtrack occupies a fixed-length memory, the backtrack will contain exactly $m$ indices, hence $k_i\,|\,m_s<i\le m$ will be unspecified.
	\item During backtracking, when reading the cell $(i,j)$ with backtrack $(r_s,(k_1,k_2,\ldots,k_m))$, given $r_s$, we recover $s=T(r_s)$, the sub-rule that applies. Hence we can determine $m_s$, which allows us to enqueue the subsequences $(i,k_1), (k_1,k_2), ..., (k_{m_s},j)$ for recursive backtracking. If $s$ refers to a terminal, we stop the backtracking.
	\item The backtracking can be returned to the user as a mapping table $T$ and a list of triplets $((i,j),r_s,(k_1,k_2,\ldots,k_{m_s}))$ where $(i,j)$ denotes the subsequence on which the sub-rule $T(r_s)$ has to be applied with concatenation indices $(k_1,k_2,\ldots,k_{m_s})$.
\ole
In short, we break parsers into normalized rules, the backtracking information is the subword, the sub-rule id (which rule to unfold) and a list of indices (how to unfold it).
\ule
In order to reduce the storage required by the backtracking indices, we can avoid storing fixed indices (where at least one of the two subsequences involved in a concatenation has a fixed size) and leverage the knowledge contained in $s$ to reconstruct the appropriate backtrack.

@@

% ------------------------------------------------------------------------------------------------
\subsection{Memory constraints}
We denote by \textit{device} the computational device on which the processing of the DP matrix (or of a computational block) is done and $M_D$ its memory. This can be the GPU or the FPGA internal memory. Usually the main memory is larger than device memory and can ultimately be extended by either disk or network storage. Without loss of generality, let the underlying dynamic programming matrices be of dimension $m \times n$.

>>>>

\newpage
\section{Architecture design and technical decisions} \label{architecture}
% ------------------------------------------------------------------------------------------------
\subsection{User facing language requirements} \label{user_lang}
The field of dynamic programming has been influenced in the recent years by a methodology known as Algebraic Dynamic Programming which uses a grammar and an algebra to separate between the parsing and the score computation:
\quote{The Algebraic Dynamic Programming approach (ADP) introduces a conceptual splitting of a DP algorithm into a recognition and an evaluation phase. The evaluation phase is specified by an evaluation algebra, the recognition phase by a yield grammar. Each grammar can be combined with a variety of algebras to solve different but related problems, for which heretofore DP recurrences had to be developed independently. Grammar and algebra together describe a DP algorithm on a high level of abstraction, supporting the development of ideas and the comparison of algorithms.}

@@

% ------------------------------------------------------------------------------------------------
\newpage
\subsection{Parsing grammar (ADP)} \label{adp_grammar}
In this section, we provide a concise description of the Algebraic Dynamic Programming. Would the reader be interested in the mathematical details of the notation, we encourage him to read the section 3 of \cite{adp}.

@@

% ------------------------------------------------------------------------------------------------
%\newpage
\subsection{Recurrences analysis} \label{recurrences}
\subsubsection{Dead rules elimination}
Dead rules\footnote{\textit{Rule} denotes a tabulation belonging to the grammar; both terms refer to the same concept, with the subtle difference that \textit{rule} emphasizes its grammar membership.} elimination analysis is straightforward: starting from the grammar's axiom, recursively collect all tabulations involved in the computation in $R$ (set of rules that are mandatory in the grammar). The dead rules $D=S \setminus R$ (where $S$ is the set of all tabulations) can safely be removed from the grammar rules. Although seemingly useless for the Scala implementation, this step is necessary to maintain coherency between Scala and CUDA rules numbering (that happen in a later stage on the valid rules). In CUDA, this analysis not only provides dead code elimination, but it also prevents useless computation execution, since all rules are computed sequentially for a particular subsequence before the next subsequence is processed.

\subsubsection{Yield analysis} \label{yield_analysis}
Since the original ADP introduces many concatenation combinators\footnote{ADP's original combinators are: $\sim\sim, \sim\sim*, *\sim\sim, *\sim*, -\sim\sim, \sim\sim-, +\sim\sim, \sim\sim+, +\sim+$.} to differentiate empty/non-empty, and floating/fixed-length concatenations, it is quite involved for the programmer to make sure that the concatenation operators exactly match the size of each pair of subsequence involved. Additionally the priority of operators varies in Scala, depending the operator's first character whereas it is possible to specify arbitrary priorities in Haskell. To overcome these issues, we propose to automate the computation of the minimum/maximum length of (subsequences) parsers.

@@

% ------------------------------------------------------------------------------------------------
\subsection{Backtracking} \label{backtracking}
In order to produce an efficient transformation from ADP-like problem description to plain C recurrences, we need to construct bottom-up recurrences from top-down parser rules. To do that, we slightly need to modify the ADP parsers in order to separate the backtracking and the scoring, because we want to obtain an efficient algorithm: backtrack writes are in $O(n^2)$ whereas score reads are proportional to the algorithmic complexity ($O(n^3)$ or more for non-serial). To deal with this problem, we are facing two options:\ul
\item \textbf{Explicit backtracking:} requires clear syntactical separation between the score and the backtrack which is not implemented in ADP, unless we consider the whole backtrack being part of the scoring (which has a big performance impact and non-constant memory requirement issues that make such GPU implementation hard and not desirable). Additionally, since the backtracking data is user-defined, there is no way to generate the backtracking algorithm automatically, hence the user also needs to provide it.
\item \textbf{Implicit backtracking:} implies that every rule needs to be normalized, and transformed such that given a rule identifier and a set of indices (subproblems breaking), it is possible to retrieve the sub-solutions combination that contribute to the problem solution. To do that we need to apply the following transformations\ol
	\item Normalize rules and identify them uniquely by exploding alternatives: each rule is decomposed into the union of multiple sub-rules uniquely identified by an index, where sub-rules do not contain alternatives (Or parsers). Let $s$ a subrule and $r_s$ its identifier, we also establish a mapping $T$ from identifier to subrule: $(r_s \to s) \in T$.
	\item Let ${\rm cc}(r_s)$ the number of concatenation contained in the sub-rule $r_s$. The data element corresponding to a rule is a pair (score, backtrack) and is named after the tabulation.\ul
		\item The score part consists of a user-defined type (a composite of primitive types, case classes and tuples)
		\item The backtrack part is a tuple $(r_s,(k_1, k_2, \ldots , k_m))$ where $m$ is the maximal number of concatenations occurring in the enclosing rule of $r_s$; more formally $m=\max_z\big[{\rm cc}(r_z) | r_z \in {\rm rule}(r_s)\big]$, and let $m_s = {\rm cc}(r_s) \le m$.
	\ule
	\item During the matrix computation of cell $(i,j)$, if the sub-rule $r_s$ applies, the backtrack will be set as $(r_s,(k_1,k_2,\ldots,k_{m_s}))$; with $i\le k_1\le k_2\le \ldots k_{m_s}\le j$. Note that if the backtrack occupies a fixed-length memory, the backtrack will contain exactly $m$ indices, hence $k_i\,|\,m_s<i\le m$ will be unspecified.
	\item During backtracking, when reading the cell $(i,j)$ with backtrack $(r_s,(k_1,k_2,\ldots,k_m))$, given $r_s$, we recover $s=T(r_s)$, the sub-rule that applies. Hence we can determine $m_s$, which allows us to enqueue the subsequences $(i,k_1), (k_1,k_2), ..., (k_{m_s},j)$ for recursive backtracking. If $s$ refers to a terminal, we stop the backtracking.
	\item The backtracking can be returned to the user as a mapping table $T$ and a list of triplets $((i,j),r_s,(k_1,k_2,\ldots,k_{m_s}))$ where $(i,j)$ denotes the subsequence on which the sub-rule $T(r_s)$ has to be applied with concatenation indices $(k_1,k_2,\ldots,k_{m_s})$.
\ole
In short, we break parsers into normalized rules, the backtracking information is the subword, the sub-rule id (which rule to unfold) and a list of indices (how to unfold it).
\ule
In order to reduce the storage required by the backtracking indices, we can avoid storing fixed indices (where at least one of the two subsequences involved in a concatenation has a fixed size) and leverage the knowledge contained in $s$ to reconstruct the appropriate backtrack.

@@

% ------------------------------------------------------------------------------------------------
\subsection{Memory constraints} \label{memory_constr}
We denote by \textit{device} the computational device on which the processing of the DP matrix (or of a computational block) is done and $M_D$ its memory. This can be the GPU or the FPGA internal memory. Usually the main memory is larger than device memory and can ultimately be extended by either disk or network storage. Without loss of generality, let the underlying dynamic programming matrices be of dimension $m \times n$.

### 2013-01-14 docs/report/introduction.tex
Nits in librna.
Rework of intro.

\newpage
\section{Introduction}
% ------------------------------------------------------------------------------------------------
\subsection{Dynamic programming}
Dynamic programming consists of solving a problem by reusing subproblems solutions. A famous example of dynamic programming is the Fibonacci series that is defined by the recurrence
\[F(n+1) = F(n)+F(n-1) \qquad \text{ with } F(0)=F(1)=1 \]
which expands to (first 21 numbers)
\[1, 1, 2, 3, 5, 8, 13, 21, 34, 55, 89, 144, 233, 377, 610, 987, 1597, 2584, 4181, 6765, 10946, ...\]

@@

% ------------------------------------------------------------------------------------------------
\newpage
\subsection{Scala and LMS}
\subsubsection{Scala}
\textit{<<Scala is a general purpose programming language designed to express common programming patterns in a concise, elegant, and type-safe way. It smoothly integrates features of object-oriented and functional languages, enabling programmers to be more productive. Many companies depending on Java for business critical applications are turning to Scala to boost their development productivity, applications scalability and overall reliability.>>}\footnote{\url{http://www.scala-lang.org}}

As the Scala \cite{scala} programming language is initially developed by the LAMP, it seems natural to use it as host language for our project, however, we would list some of its features \cite{scala_api} that makes it an interesting development language for this project:\ul
\item The functional programming style and syntactic sugar offered by Scala allows concise writing of implementation, analysis and transformations of our DSL, which would have been much more complex and tiresome in an imperative language like C.
\item Scala is largely adopted in the industry, which makes both the adoption of related project easier and offer a steeper learning curve to their potential users.
\item Finally, through the Java VM and JNI interface, Scala offers the possibility to load dynamically external libraries, to leverage best underlying hardware by mixing with CUDA kernels to obtain optimal performance.
\ule

\subsubsection{Lightweight Modular Staging}
Lightweight Modular Staging (LMS) \cite{lms}, \cite{lms_thesis} is a runtime code generation built on top of Scala virtualized \cite{scala_virtualized} that uses types to distinguish between binding time (compilation and runtime) for code compilation. This framework has been extended to generate from the same source code efficient implementation for heterogeneous platforms at runtime (Delite) \cite{lms2}, \cite{lms3}, \cite{delite}.

 Through extensive use of component technology, lightweight modular staging makes an optimizing compiler framework available at the library level, allowing programmers to tightly integrate domain-specific abstractions and optimizations into the generation process.

LMS can be leveraged to transform Scala code into its C-like equivalent. However, the concern in this project is that the code for the GPU would be sensibly different from the original CPU code as both implementation serve different purposes: CPU version (Scala) is more general whereas the GPU version trades some functionalities for performance and suffer additional restrictions, in particular for memory management and alignment. Hence the use of LMS would be restricted to user-specific function, over which our DSL has no control.

>>>>

% abstract = resume the paper
% introduction
%

% Paper introduction:
% Problem to solve, what exists (related work) and how do we compare to other
% Contributions (3): 3 tensed sentences
% Benchmarks => evaluation metrics, prove introduction statements by evaluation
\subsection*{Abstract}
Dynamic programming is a common pattern of Computer Science to solve problem that follow the Bellman's principle: optimal solutions depends on optimal sub-problem solutions. Solving a dynamic programming problem decomposes in two phases: filling one or more matrices with sub-problems intermediate results and decomposing how the final result was constructed (backtracking). Yet matrices computation formulae and recurrences might be difficult to express, indexing element is often error prone and writing an efficient parallel implementation can take a significant amount of time. In this project, we present \textit{DynaProg}, a language embedded in Scala (DSL) to address dynamic programming problems on heterogeneous platforms. DynaProg allows the programmer to write concise programs based on a pair of parsing grammar and algebra; these program can then be executed either on CPU or on GPU.

We evaluate the performance of our implementation against existing work and own ad-hoc implementations for both the CPU and GPU versions of DynaProg.

Experimental results show an average speedup of {\color{red} XXX} for large problems when they are run on the GPU (compared to CPU). Compared to ad-hoc GPU implementations, the generated parsers have an average slowdown of {\color{red} XXX}.

The CPU implementation has a slowdown of {\color{red} XXX} for {\color{red} PROBLEM} against {\color{red} REF,GAPC?}.
The GPU implementation has a slowdown of {\color{red} XXX} for {\color{red} PROBLEM} against {\color{red} REF,CudAlign, RNAFold}.

\vfill
This project has been achieved in collaboration with Manohar Jonnalagedda. I also would like to thank the LAMP team, including Eugene Burmako, Sandro Stucki, Vojin Jovanovic and Tiark Rompf who provided insightful advices and suggestions. I hope you will enjoy your reading. \vspace{.3cm}\\
\textit{Thierry Coppey}

% ------------------------------------------------------------------------------------------------
\newpage
\setcounter{tocdepth}{2} \tableofcontents
\newpage
\section{Introduction}
Dynamic programming (DP) is a technique used to solve combinatorial optimization problems that verify the Bellman's principle: optimal solutions depends on optimal solutions of sub-problem. Dynamic programming often allows to find solutions in an exponential search space (number of solutions) in polynomial running time. The user is usually interested in one optimal solution of the problem, but he might also desire to have \textit{all} optimal solutions (co-optimal), a fixed number of near-optimal solutions, or some synthetic properties of the search space such as its size or the sum of all scores.

Dynamic programming problems arise in several areas of applied Computer Science such as biosequence analysis, natural language processing and operational research. Unfortunately, they appear in multiple variations, and often with a considerable degree of sophistication such that there is a mismatch between textbook solutions and the real-world implementation. Additionally, implementation debugging is tedious and require a lot of time, and little changes in the formulae might imply large rewrites of the matrices and recurrences. \cite{gapc_yield}

For efficiency reasons, the process of dynamic programming computation is split in two: first intermediate scores are memoized in a matrix (tabulation), and reused to construct scores for larger problems; then a backtrack stage retrieve the solution associated with the optimal score for the problem. This solution describes how to obtain the optimal score and is called trace or backtrack trace, and heavily depends on the matrix design. The obtained information then needs to be presented in a form that makes sense for the user (or possibly might be used to drive further computations). Each of these two phases must be tightly correlated as a divergence would be a source of errors; would the recurrence formulae be even slightly changed, the programmer must ensure that both phase are still synchronized, which might be tedious.

Finally, once the implementation is correct, it is possible to turn it into an efficient implementation for specific architectures such as multi-CPU, GPU or programmable hardware (FPGA). However, the domain specialist who write the recurrences might not be very familiar with these platforms, whereas parallelization and hardware experts might not deeply understand the domain of the dynamic programming recurrences.

To solve these concerns, Algebraic Dynamic Programming (ADP) \cite{adp} proposes a language-independent declarative approach that separate the concerns of dynamic programming algorithms into four distinct components that are tightly connected:\ol
\item The search space is described using a context-free \textbf{parsing grammar} that describes how to construct intermediate candidates whose score might be inserted in the matrix.
\item Constructed candidates are then evaluated by a \textbf{scoring function} (where all these functions form an \textbf{algebra}), so that they can be compared appropriately.
\item The optimization objective (which candidates to retain as possible solutions) is described using an \textbf{aggregation function} that operates on the scores previously obtained.
\item Finally, results are \textbf{tabulated} (memoized in an array) in the corresponding matrices. The tabulation regulates the trade-off between running time and space efficiency by memoizing appropriate results that are reused multiple times.
\ole

By the use of a parsing grammar, ADP makes the candidate structure explicit. An additional signature serves as interface between the grammar, the scoring algebra and the aggregation function which makes possible that different grammars share different algebra or vice versa. Tabulation indices issues are hidden from the programmer, hereby removing one source of errors.

Finally, since the expression of the dynamic program is formalized and abstracted into a grammar and algebra, it becomes possible to convert it to efficient recurrences for many-core platforms such as GPUs. \cite{adp_gpu}

% -----------------------------
DynaProg implements the concepts of ADP in Scala as an embedded DSL (domain-specific language) with a syntax matching similar to the combinators parsers of Scala library\footnote{See \url{http://www.scala-lang.org/api/current/index.html\#scala.util.parsing.combinator.Parsers}}. It extends over original ADP by allowing grammars for pairing two sequences (multi-track grammars), simplifies the process of writing program by inferring additional informations (see \ref{yield_analysis}) and can translate them into efficient CUDA\footnote{Compute Unified Device Architecture: a parallel computing platform and programming model created by NVIDIA, supported by their graphics processing units (GPUs).} program that are competitive to their handwritten counterpart. Since the program is formalized, it can be analyzed to remove unused grammar productions (dead code elimination) and avoid some non-termination issues; since it is generated, correct scheduling is guaranteed and indices errors are avoided, hereby producing an arguably more reliable program.

DynaProg provides a generic way of backtracking the results, such that the same backtrack trace can be used with multiple algebras if they share the same grammar. This allows to construct a two steps pattern for solving problems: first DP problem is solved, then from optimal solution backtrack, the desired result is computed. For example, assume the problem of multiplying a chain\footnote{Assuming matrices are of appropriated dimension to be multiplied with each other} of matrices efficiently. In the first step, optimal execution scheduling (or parenthesization) is found using dynamic programming. At the second step, the backtrack trace of optimal solution is used to multiply the actual matrices (given the corresponding algebra).

Finally, offloading dynamic programming computations to CUDA devices has been made effortless for the programmer: it suffices to enable code generation to schedule dynamic compilation and execution of the GPU-optimized program, as if it was executed in plain Scala.

This project resulted is an open-source\footnote{\url{https://github.com/manojo/lamp-dp-mt}} implementation of dynamic programming parsers for Scala optimized for both on CPU and GPU and featuring several analysis (see section~\ref{architecture}) to ease the writing of dynamic programs. Its contribution is an automated approach to encode and process backtracking information such that reconstruction complexity is reduced and backtrack trace be exchanged among different algebras sharing the same grammar.

The rest of the document consists of:\ul
\item A brief description of dynamic programming background (\ref{intro_dp}), followed by an explanation of some of the Scala programming language (\ref{intro_scala}) and LMS framework (\ref{intro_lms}) features.
\item Section~\ref{problems} proposes a classification of DP problems in terms of matrix shape and dependencies. A detailed analysis of some specific problems is provided. Related work addressing the dynamic programming challenges is then presented (\ref{related_work}).
\item In section~\ref{architecture} we describe the whole parser stack in an abstract manner, going from the user facing language (\ref{user_lang}, \ref{adp_grammar}) to optimizations (\ref{recurrences}, \ref{backtracking}) and implementation constraints (\ref{normalization}, \ref{memory_constr}), describing all the architectural decisions we made.
\item Section~\ref{implementation} describes how these idea are concretely implemented in the form of a DSL for Scala (\ref{scala_parsers}) and in particular how is efficient CUDA code generated (\ref{codegen})
\item Finally, in section~\ref{benchmarks} we evaluate the performance of our work by providing appropriate benchmarks against existing implementations
\ule

%\item Propose a systematic approach to encode backtracking information such that the backtracking process can be made linear to the size of the problem
%\item Provide an concrete implementation in the form of a language embedded DSL in Scala, leveraging the grammar and algebra concepts of ADP
%\item Describe two implementations: Scala for CPU (focusing on multiple backtracking) and an CUDA for GPU (focusing on efficiency)

%The contributions of this project are: \ul
%\item A classification of dynamic programming problems characteristics in terms of matrix shape and recurrence formulae dependencies.
%\item A systematic approach to convert a top-down recurrence description (grammar) into efficient bottom-up 
%\item A systematic approach to process backtracking information (focus on running time and memory efficiency)
%\item A language embedded in Scala (DSL) to express DP problems concisely (based on ADP)
%\item Two implementations: Scala for CPU (features) and an CUDA for GPU (efficiency)
%%\item Reuse of existing compiler technology (fusion) for a specific purpose
%%\item State of the art parallel implementation of these classes on GPUs
%%\item Normalization of the grammar into efficient productions
%%\item Code generator to transform a grammar into efficient code for CPU, GPU (and FPGA)
%\ule

% ------------------------------------------------------------------------------------------------
\newpage
%\section{Background}
\subsection{Dynamic programming} \label{intro_dp}
Dynamic programming consists of solving a problem by reusing subproblems solutions. A famous example of dynamic programming is the Fibonacci series that is defined by the recurrence
\[F(n+1) = F(n)+F(n-1) \qquad \text{ with } F(0)=F(1)=1 \]
which expands to (first 21 numbers)
\[1, 1, 2, 3, 5, 8, 13, 21, 34, 55, 89, 144, 233, 377, 610, 987, 1597, 2584, 4181, 6765, 10946, ...\]

@@

% ------------------------------------------------------------------------------------------------
\newpage
\subsection{Scala} \label{intro_scala}
\textit{<<Scala is a general purpose programming language designed to express common programming patterns in a concise, elegant, and type-safe way. It smoothly integrates features of object-oriented and functional languages, enabling programmers to be more productive. Many companies depending on Java for business critical applications are turning to Scala to boost their development productivity, applications scalability and overall reliability.>>}\footnote{\url{http://www.scala-lang.org}}

As the Scala \cite{scala} programming language is developed by our laboratory (LAMP, EPFL), it seems natural to use it as host language for our project, however, we would list some of its features \cite{scala_api} that makes it an interesting development language for this project:\ul
\item The functional programming style and syntactic sugar offered by Scala allows concise writing of implementation, analysis and transformations of our DSL, which would have been much more complex and tiresome in an imperative language like C.
\item Scala is largely adopted in the industry, which makes both the adoption of related project easier and offer a steeper learning curve to their potential users.
\item Finally, through the Java VM and JNI interface, Scala offers the possibility to load dynamically external libraries, to leverage best underlying hardware by mixing with CUDA kernels to obtain optimal performance.
\ule

{\color{red} talk of features: traits, ... }

% ------------------------------------------------------------------------------------------------
\subsection{Lightweight Modular Staging} \label{intro_lms}
Lightweight Modular Staging (LMS) \cite{lms}, \cite{lms_thesis} is a runtime code generation built on top of Scala virtualized \cite{scala_virtualized} that uses types to distinguish between binding time (compilation and runtime) for code compilation. This makes possible to annotate parts of the code with special types, such that their compilation is delayed until the program is executed. At run time, these parts are represented as multiple nodes that serve as the basis for another compilation phase where all the code executed until this point can provide additional information to produce a more efficient compilation. The process of delaying the compilation is known as \textit{lifting} whereas \textit{lowering} corresponds to transforming this intermediate representation into executable code. Through extensive use of component technology, lightweight modular staging makes an optimizing compiler framework available at the library level, allowing programmers to tightly integrate domain-specific abstractions and optimizations into the generation process.

{\color{red} move delite to related work ... }
LMS code generation is not limited to Scala, it can also target other languages. To automate code generation and execution flow, Delite \cite{lms2}, \cite{lms3}, \cite{delite} leverages LMS to generate from the same source code efficient implementation for heterogeneous platforms at runtime. This can be used to transform operations on collections into efficient parallel GPU implementation.

At first glance, LMS seems the ideal candidate to transform Scala code into its C-like equivalent. However, the concern in this project is that the GPU code sensibly differers from the original CPU code because the two implementations serve different purposes: CPU version (Scala) is more general whereas the GPU version trades some functionalities for performance and suffer additional restrictions, in particular for memory management and alignment. Ad-hoc C translation seems more appropriate than writing Scala code and convert it with LMS because LMS only understand a subset of both languages.

LMS would still be helpful for generating user-specific functions, as they are independent of the rest of the program and user wants to write his functions only once for both Scala and CUDA.

### 2013-01-14 docs/report/problems.tex
Rework of intro.

% ------------------------------------------------------------------------------------------------
\newpage
\subsection{Related work}
Work related to dynamic programming can be separated in two categories: ad-hoc implementations and grammar-based implementations. The former focus on the performance for a specific problem whereas the latter generalize and formalize the dynamic programming problem description into a parsing grammar paired with a costing algebra.

>>>>

% ------------------------------------------------------------------------------------------------
\newpage
\subsection{Related work} \label{related_work}
Work related to dynamic programming can be separated in two categories: ad-hoc implementations and grammar-based implementations. The former focus on the performance for a specific problem whereas the latter generalize and formalize the dynamic programming problem description into a parsing grammar paired with a costing algebra.

### 2013-01-14 docs/report/report.tex
Nits in librna.
Rework of intro.

\title{DynaProg for Scala}
\subtitle{A Scala DSL for Dynamic Programming on CPU and GPU}
\begin{document}
\maketitle
\input{introduction}
\input{problems}
\input{design}
\input{implementation}
\section{Benchmarks}
In an attempt to provide realistic benchmarks, we tried to gather related implementations. The authors of \cite{gpu_atlp} did not respond to our multiple solicitations. The authors of \cite{swat_mega} were very friendly and provided us their source code. Unfortunately, since they address a different category of problem (they focus on huge serial problems whereas we focus on smaller non-serial problems) their implementation might be biased towards large sequences that our implementation cannot address. Finally, we asked lately the authors of \cite{gpu_rnafold} {\color{red} who did not respond so far}.

@@

Candidates : 
- CUDASW++
- SeqNFind

{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ 
XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}

@@

\subsubsection{Test 2 : CUDAlign vs CUDA parsers for small sequences}

>>>>

\title{DynaProg for Scala}
\subtitle{A Scala DSL for Dynamic Programming on CPU and GPU}
\begin{document}
\maketitle
\input{introduction}
\input{problems}
\input{design}
\input{implementation}
\section{Benchmarks} \label{benchmarks}
In an attempt to provide realistic benchmarks, we tried to gather related implementations. The authors of \cite{gpu_atlp} did not respond to our multiple solicitations. The authors of \cite{swat_mega} were very friendly and provided us their source code. Unfortunately, since they address a different category of problem (they focus on huge serial problems whereas we focus on smaller non-serial problems) their implementation might be biased towards large sequences that our implementation cannot address. Finally, we asked lately the authors of \cite{gpu_rnafold} {\color{red} who did not respond so far}.

@@

% ------------------------------------------------------------------------------------------------
{\center\color{red} \noindent\rule{16cm}{0.4pt} \\ XXX: CONTINUE HERE :XXX \\}
% ------------------------------------------------------------------------------------------------

@@

\subsubsection{Test 2 : CUDAlign vs CUDA parsers for small sequences}
Also CUDASW++
